\documentclass[twocolumn,12pt]{article}
\usepackage{tikz}
\usetikzlibrary{calc}
\newcommand{\tikzmark}[1]{\tikz[overlay,remember picture] \node (#1) {};}
\newcommand{\underarrow}[2] {
  \begin{tikzpicture}[overlay,remember picture,out=340,in=210,distance=0.3cm]
    \draw [->,shorten >=3pt,shorten <=-3pt] ({#1}.south) to ({#2}.west);
  \end{tikzpicture}
}  
\newcommand{\Ex}{\mathbb{E}}
\usepackage{graphicx}
\usepackage{dblfloatfix}
\usepackage[superscript,biblabel]{cite}
\usepackage{multirow,tabularx}
\usepackage{hhline}
\usepackage{amsmath}

\newcommand\indep{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}

\title{Inferring Causality from Finite Data using Conditional Independence}
\author{Daniel Speyer}
%\address{dls2192@columbia.edu}
\begin{document}

\maketitle

\section{Introduction}

Inferring causal graphs from observational data is a widely-sought
goal in statistics.  The most common tool for it is conditional
independence.  The specifics of a causal graph determine which node
are independent conditioned on which others by the ``bayes ball''
rule, and therefore it should be possible to observe the independences
and work backward to the causal graph.

This has proved to be more difficult in practice.  In particular,
observing independence is not as simple as it sounds.  Pearl suggests
we test joint conditional probability distributions for equality\cite{Pearl}.
Leaving aside the curse of dimensionality, this assumes we have the
exact distributions.  If all we have is a finite random sample from
those distributions, we can construct posteriors, but we cannot perform
an equality test.  Spirtes, Glymour and Shine  encourage us to assume
``the statistical decisions required by the [causal inference]
algorithms are correct
for the population'', while admitting that this ``is often not met in
practice''\cite{Spirtes}.  Others have described this as ``possessing an
independence oracle''\cite{Peters,pcalg}.  Many simply include a ``test if $a\indep
b|s$'' step in their algorithms, assuming the reader will already know
how.

In practice, the usual solution is to treat independence as a null
hypothesis, try to reject it at some p threshold, and treat any
failure as establishing it.  Needless to say, this is incorrect.

Dealing with finite data means the possibility of dealing with too
little data.  The elegant solution is to give some numerical
expression of confidence which becomes ``I don't know'' when the data
gets too small.  This solution has another benefit: in the biomedical
context, it is routine to test thousands of equally plausible
hypotheses at once.  A flat accuracy of 99\% is unhelpful in the face
of this, but a numeric expression of confidence allows proper
compensation.

\subsection{Inspiring Problem: Crohn's Disease}

This paper is optimized around a specific practical problem:
untangling the microbiome's role in Illial Crohn's Disease.  It is
well established that there are many differences in the intestinal
bacteria of healthy people and of people with the disease\cite{hofer}, but it is
not established whether the differences of bacteria \textit{cause} the
disease.  If we can find a set of bacteria such that
p(disease|do(bacteria)) is low, we will have a cure for the disease.
There have been attempts to determine this by randomized controlled
trial, but early results are discouraging\cite{rctma} and the number
of species, combined with other relevant
variables, make exhaustive RCTs impractical.

Data is available on this problem\cite{data}, including genetic, microbiome and
health information, but for only 58 patients.  The microbiome
information is a series of 16S reads, but can generally be described
as ``present'' or ``absent'', with very low concentrations of a
species rounded off to ``absent''.  This comes much closer to fitting
the empirical distributions than any convenient scalar formula (see
figure 1) which is why this paper will use binary variables.

\section{Extending Graphs}

Let us begin with the simplest case.  Suppose we have a known cause, a
known effect, and a variable which connects to the effect in an
unknown way, that is $A \rightarrow B$ -- $C$.  We do not observe a
correlation between $A$ and $C$, but that might only mean our test is
underpowered.  For Crohn's Disease, $A$ would be mutations in the NOD2
gene, $B$ would be the disease, and $C$ would be each of [number]
species of bacteria.

For now, we will assume that there is no direct causal
link between $A$ and $C$, and furthermore that there are no unobserved
confounders or selection effects.  We will consider these later.
This leaves us only two models, $A \rightarrow B \rightarrow C$ or $A
\rightarrow B \leftarrow C$.  We can call these ``chain'' and
``collide'' models, or $m_{ch}$ and $m_{co}$ for short.  Can we
distinguish between them?

Yes.  Let us consider
$p(A,C)$:

\begin{eqnarray*}
p(A,C|m_{ch}) & = & \sum_B p(A)p(B|A)p(C|B) \\
p(A,C|m_{co}) & = & p(A)p(C)
\end{eqnarray*}

We do not actually know the terms on the right side of those
equations, so let us parameterize both models with $\theta$ and
rewrite the equations as:


\begin{multline*}
  p(A,C|m_{ch}) = \\
  \int \left ( \sum_B p(A|\theta) p(B|A,\theta) 
  p(C|B,\theta) \right ) p(\theta) d\theta \\
  p(A,C|m_{co}) = \int p(A|\theta)p(C|\theta)p(\theta) d\theta
\end{multline*}

We can learn $p(\theta)$ from available data using dirichlet priors.
The integrals would be difficult algebraically, but they can be
adequately approximated with monte-carlo sampling.

Once we have these, let $n_{a,c}$ be the count of datapoints with
A=a,C=c and we can use:

\begin{equation*}
p(n_*|m) \propto \prod_{a,c} p(a,c|m)^{n_{a,c}}
\end{equation*}

Proportional instead of equal because there is a combinatoric term,
but it cancels when taking odds ratios so it can be safely ignored.

From here, we can apply standard bayesian updating.  Empirically, this
works well, in a test of 10k runs of each model with random parameters
(see figure).

\subsection{Non-Simply-Connected Graphs}

What if there is a causal effect $A\rightarrow C$?  It must be weak
enough not to be detected, but that doesn't say much.  Let us consider
it by cases.

\subsubsection{False Colliders}

Can a true graph of
$A\tikzmark{a}\rightarrow B \rightarrow{C}\tikzmark{c}$
\underarrow{a}{c}
produce a $p(A,C)$ more similar to a graph
$A \rightarrow B \leftarrow C$?  This would require the direct and
indirect influence of $A$ on $C$ to cancel out rather precisely.  If
the indirect influence is stronger, we will pick the correct model,
albeit underconfidently.  If the direct influence is too strong, we
will see a correlation between $A$ and $C$.  And if the direct
influence is in the same direction as the indirect, the correct model
will remain the better fit (though both models will be worse).

To put this more precisely, a false collider occurs when:

\begin{equation*}
  \prod_{A,C} p(A,C|m_{co})^{n_{A,C}}  > 
  \prod_{A,C} p(A,C|m_{ch})^{n_{A,C}}
\end{equation*}
\begin{equation*}
  MI(A,C)  <  MI(B,C)
\end{equation*}

Which is difficult to work with algebraically, but can be reasonably
approximated as:

\begin{equation*}
  \sum_{A,C} \log \left ( \frac{p(A,C|m_{co})}{p(A,C|m_{ch})} \right )
  p(A,C|m_{NSC})  >  0
\end{equation*}
\begin{equation*}
  |p(C|a)-p(C|\bar{a})| < |p(C|b)-p(C|\bar{b})
\end{equation*}

Which is still difficult to work with, but easy to examine
numerically.  Assuming all probability are equally likely, roughly
49\% of causal networks are never considered because $A\not\indep C$
is more apparent than $B\not\indep C$, 28\% give the proper direction
and 23\% give false colliders.

\subsubsection{False Chains}

Similarly, can a true graph of
$A\tikzmark{a2}\rightarrow B \leftarrow{C}\tikzmark{c2}$
\underarrow{a2}{c2}
produce a $p(A,C)$ more similar to a graph
$A \rightarrow B \rightarrow C$?  Again, this requires rather precise
parameters.

\section{Parsimony}

Given 

\begin{thebibliography}{11}
\bibitem{Pearl} Pearl ``Causality: Models, Reasoning and Inference''
  Chapter 2 \textit{Springer} 2008

\bibitem{Spirtes} Spirtes, Glymour \& Scheines ``Causation,
  Prediction, and Search'' Section 5.4 \textit{MIT Press} 2001

\bibitem{Peters} Peters, Mooij, Janzing \& Sch{\"o}lkopf ``Causal
  Discovery with Continuous Additive Noise Models'' \textit{Journal of
    Machine Learning Research} 2015

\bibitem{pcalg} Kalisch, M{\"a}chler, Colombo, Maathuis \&
  B{\"u}hlmann ``Causal Inference Using Graphical Models with the R
  Package pcalg'' \textit{Journal of Statistical Software} 2012

\bibitem{hofer} Hofer ``Bacterial imbalance in Crohn's disease''
  \textit{Nature Reviews Microbiology} 2014

\bibitem{rctma} C Prantera ``Probiotics for Crohn’s Disease: What
  Have We Learned?'' \textit{Gut} 55.6(2006):757–759, 2016

\bibitem{data} E Li, D Frank \& R Sartor, ``Effect of Crohn’s Disease Risk
  Alleles on Enteric Microbiota'' \textit{Encyclopedia of
    Metagenomics},  1-5, 2014
  
\end{thebibliography}

\end{document}
